Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-06 WholePlotTrajectories.Rmd
discuss

Disturbance clearly enhanced community taxonomic richness until an intensity threshold (20-25% of initial AGB loss), while the taxonomic evenness was somewhat decoupled from the disturbance intensity, as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The IDH well predicted the response of taxonomic richness to disturbance, and to a lesser extent the taxonomic evenness.
Post-disturbance communities combine pre-disturbance surviving trees and recruited trees, either from the remaining pre-disturbance community or from a disturbance-specific community.
In the Guiana Shield, the pool of true pioneer species specifically recruited after disturbance is restricted to a few common genera (e.g. Cecropia spp., Vismia spp.) 
Below the intensity threshold the taxonomic richness of the pre-disturbance community is maintained and the recruitment of pioneers, previously infrequent or absent, increases both taxonomic richness and evenness.
The disturbance intensity determined the relative importance of pioneers recruited, so the taxonomic richness and evennes increase all the more so that disturbance is intense.
However, disturbance in the same time decreases the taxonomic richness of the pre-disturbance community, and beyond the intensity threshold the decrease exceeds the taxonomic enrichment from pioneers.
The taxonomic evenness for its part is less sensitive to the loss of rare species and the disturbance impact is mitigated as pioneers balance the hyper-dominance of undisturbed tropical forests.

Conversely the IDH was disproved for the functional diversity.
Pioneer species are functionally highly different, and whenever the disturbance intensity their increased abundance suffice to increase the community functional diversity and evenness.

Following disturbance the taxonomic richness droped at first place independently of the disturbance intensity, but then increased through a species gain that was dermined by the disturbance intensity.
Below the intensity threshold the establishment of long-lived pioneers significantly enhanced both taxonomic and functional richness and evenness in the long-term.
Beyond the intensity threshold the establishment of short-lived pioneers rapidly filling the environmental space made available did not increased the taxonomic richness and entailed a decrease in the functional diversity after 10 years.
Likely, after intense disturbance the long-lived pioneers will succeed and post-disturbance trajectories will match those of intermediate disturbance, which explains the decoupling in time between high and low disturbance intensity trajectories.

Species first established after disturbance remained in the community but became less dominant, replaced the abundant pre-disturbance species, but the recovery of rare species was hampered in the long-term [@Hubbell2001;@Chave2004].

@@

The recovery of the most infrequent species followed the lottery recruitment rules [@Busing2002] and was hampered by the increasing competition among species, implying long term recovery [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Following disturbance the space and resources made available corresponded to a rapid decrease of the functional redundancy within the initial functional space.
The high light, space and nutrient niches were rapidly filled by species, mainly the most dominant and frequent ones.
Thereafter competitive exclusion emerged following the filling of the niches and limited the recruitment of infrequent species [@Busing2002].
The recovery of communities taxonomic richness and evenness relied upon the recovery of the initial functional redudancy.
Although underway 30 years after disturbance the recovery remained unachieved and after 30 years the recovery it was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific communities, with less species and more pioneers [@Haddad2008; @Burslem2000; @Martin2013].
Besides the slowed recovery of rare species increased the risks to loose cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species mighthave unique functions i nthe ecosystem or be a key for some fauna.

>>>>

Disturbance clearly enhanced community taxonomic richness until an intensity threshold (20-25% of initial AGB loss), while the taxonomic evenness was somewhat decoupled from the disturbance intensity, as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The IDH well predicted the response of taxonomic richness to disturbance, and to a lesser extent the taxonomic evenness.
Post-disturbance communities combine pre-disturbance surviving trees and recruited trees, either from the remaining pre-disturbance community or from a disturbance-specific community.
In the Guiana Shield, the pool of true pioneer species specifically recruited after disturbance is restricted to a few common genera (e.g. Cecropia spp., Vismia spp.) 
Below the intensity threshold the taxonomic richness of the pre-disturbance community is maintained and the recruitment of pioneers, previously infrequent or absent, increases both taxonomic richness and evenness.
The disturbance intensity determined the relative importance of pioneers recruited, so the taxonomic richness and evennes increase all the more so that disturbance is intense.
However, disturbance in the same time decreases the taxonomic richness of the pre-disturbance community, and beyond the intensity threshold the decrease exceeds the taxonomic enrichment from pioneers.
The taxonomic evenness for its part is less sensitive to the loss of rare species and the disturbance impact is mitigated as pioneers balance the hyper-dominance of undisturbed tropical forests.
Conversely the IDH was disproved for the functional diversity.
The pioneer species were functionally highly different from the pre-disturbance community composition, so their recruitment increased the community functional richness and eveness after disturbance whenever its intensity.

Community post-disturbance trajectories showed a first drop of the taxonomic richness, similar among communities, followed by a species gain determined by the disturbance intensity.
Until an intensity threshold the gain in taxonomic richness increased with the disturbance intensity.
This corresponded to the establishment of long-lived pioneers enhancing community taxonomic richness and evenness in the long term.
Beyond the intensity threshold a few short-lived pioneers occupied the vacated environmental space, preventing the establishment of further species. 
The establishment of short-lived pioneers, functionally very different from the pre-disturbance community, rapidly and significantly increased the functional diversity but it started decreasing after 10 years.
Below the disturbance intensity threshold on the other hand the establishment of long-lived pionneers entailed a later and lower but long-lasting increase of the functional diversity.
Likely the establishment of long-lasting pioneers will follow the decline of short-live pionneers and we expect the taxonomic and functoinal trajectories to catch up with those observed after intermediate disturbance.

@@

The functional redundancy within the pre-disturbance functional space decreased for 15 years because of the loss of species, and was not compensated by the functionally different pionneers.
The depth of the redundancy decrease relied on the disturbance intensity that enhanced the initial decrease and the functional differences of pionneers from the pre-disturance composition.
After 15 years the first established pioneers pionneers started to be replaced by more competitive long-lived pionneers or late-successional species resembling more the pre-disturbance functional composition and hence increasing the functional redundancy.
This replacement though is stochastic and follows the lottery recruitment so because of increasing competition it is increasingly difficult recover pre-disturbance species,specifically the most infrequent ones, as the functional redundancy recovers [@Busing2002].
The recovery time proved long as for some community it was just initiated after 30 years but it is thus difficult to estimate [@Trenbath1999;@Elmqvist2003;@Diaz2005].

The long-term impact of disturbance on community functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific communities, with lower taxonomic richness and more pioneers [@Haddad2008; @Burslem2000; @Martin2013].
Besides the slowed recovery of rare species increased the risks to loose cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species mighthave unique functions i nthe ecosystem or be a key for some fauna.

### 2018-07-06 WholePlotTrajectories.tex
discuss

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) removed. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) questioned the
recovery of communities taxonomic and functional characteristics and
identified the underlying assembly processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and elucidated
its translation into different trajectories, and (iii) questioned
community recovery time.

@@

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB removed. They were
besides analysed through polynomial regression between (i) taxonomic and
functional richness, evenness and diversity and (ii) the initial \%AGB
removed at 10, 20 and 30 years after disturbance.

@@

Among polynomial regressions between (i) the \%AGB removed and (ii)
taxonomic richness, evenness and functional diversity after 10, 20 and
30 years the second degree better predicted the hump-shaped curve of the
disturbance impact in function of the intensity \ref{fig:IDHplot}. The
correlation with disturbance intensity was weak for the taxonomic
richness (\(R^2<0.37\)) and taxonomic evenness (\(R^2<0.43\)) but highly
significant for functional diversity (\(0.60<R^2<0.81\)).

@@

\caption{Relationship between the initial \%AGB removed and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)} and functional diversity \textbf{(c)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

Communities taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years. The taxonomic differences
among communities, marked before disturbance by the distinct starting
points on the NMDS axis 2, were maintained throughout recovery
trajectories. More than commonly thought, post-disturbance trajectories
depended on community initial composition, that partly determined the
pool of recruited species and constrained the trajectories towards the
initial composition. The high resilience of communities taxonomy
revealed that species not belonging to the pre-disturbance community
were hardly recruited because of the commonness of dispersal limitation
among tropical tree species (Svenning, Wright JECOL 2005).

Community functional composition followed similar trajectories that were
not so distinct in the functional space\ldots{}\ldots{}. Because the
pre-disturbance survivor trees mirror the initial communities
\citep{Herault2018}, the functional changes relied upon recruited trees
and the enhanced growth and survival of previously infrequent species
and functional types. Disturbance vacated environmental niches of high
light, space and nutrient availability, filled by competitive pioneers
that became dominant and determined the functional characteristics of
the community \citep{Grime1998}. Before disturbance and over communities
recovery time though, these species would be excluded by long-lived,
more resistant and shade tolerant species. The recovery of communities
functional characteristics, that mostly rely on dominant species
according to the ``vegetation quantity effect'' \citep{Grime1998},
suggested a similar recovery of the dominant species \citep{Molino2001}.
significant functional shifts towards resource-acquisitive strategies
(sharp increase in the SLA, leaf thickness and bark thickness and
decrease in wood specific gravity, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  IDH in Space (Disturbance gradient) The IDH consistently (predicted)
  the response of functional diversity to disturbance, but the theory
  was disproved regarding communities taxonomic richness and evenness.
  The functional range of species that could establish after disturbance
  was determined by the expanse of the environmental space made
  available by disturbance.
\end{enumerate}

In contrast the IDH translated in a decrease of the taxonomic richness
above an intensity threshold while the taxonomic evenness was decoupled
from disturbance intensity, as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The taxonomic richness and evenness of recruited
species was fixed and at community scale the post-disturbance taxonomic
characterisitcs depended on the combination of these recruits and the
remaining pre-disturbance community.

The intensity of the disturbance determined the balance between trees
recruited in the pre-disturbance community and those recruited in
disturbance-specific community, functionally more diverse than
undisturbed forests but with lower taxonomic richness and higher
evenness.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  IDH in Time For the first years following disturbance community
  functional diversity increased through the filling of the vacated
  environmental space without competitive exclusion contraints. The
  maximum functional diversity was reached at the same time so the rate
  of space filling was similar among communities. The maximum was
  followed by the emergence of competitive exclusion processes bringing
  the functional diversity back to pre-disturbance levels and made the
  disturbance hypothesis intermediate in time.
\end{enumerate}

For medium disturbance intensity enhancing the taxonomic richness and
evennes, the taxonomic trajectories reached asymptots of high richness
and high evenness decoupled from their functional trajectories. Species
first established after disturbance remained in the community but became
less dominant, replaced the abundant pre-disturbance species, but the
recovery of rare species was hampered in the long-term
\citep{Hubbell2001, Chave2004}.

@@

The recovery of the most infrequent species followed the lottery
recruitment rules \citep{Busing2002} and was hampered by the increasing
competition among species, implying long term recovery
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Following disturbance the space and resources made available
corresponded to a rapid decrease of the functional redundancy within the
initial functional space. The high light, space and nutrient niches were
rapidly filled by species, mainly the most dominant and frequent ones.
Thereafter competitive exclusion emerged following the filling of the
niches and limited the recruitment of infrequent species
\citep{Busing2002}. The recovery of communities taxonomic richness and
evenness relied upon the recovery of the initial functional redudancy.
Although underway 30 years after disturbance the recovery remained
unachieved and after 30 years the recovery it was underway but remained
unachieved. This alteration of the functional redundancy meant a lower
resilience of pre-disturbance communities and higher chances to see the
persistence of disturbance-specific communities, with less species and
more pioneers \citep{Haddad2008, Burslem2000, Martin2013}. Besides the
slowed recovery of rare species increased the risks to loose cornerstone
specie, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.

>>>>

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) questioned the
recovery of communities taxonomic and functional characteristics and
identified the underlying assembly processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and elucidated
its translation into different trajectories, and (iii) questioned
community recovery time.

@@

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analysed through polynomial regression between (i) taxonomic and
functional richness, evenness and diversity and (ii) the initial \%AGB
loss at 10, 20 and 30 years after disturbance.

@@

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic richness, evenness and functional diversity after 10, 20
and 30 years best predicted the hump-shaped curve of the disturbance
impact along the disturbance intensity gradient \ref{fig:IDHplot}. Along
the disturbance intensity gradient the disturbance impact was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness and the functional diversity, the latest being almost linear.
The regression model better predicted the functional diversity
(\(0.60<R^2_{Functional}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Richness}<0.4\), and \(-0.15<R^2_{Evenness}<0.43\)
respectively)

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)} and functional diversity \textbf{(c)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

Communities taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years. The taxonomic differences
among communities, marked before disturbance by the distinct starting
points on the NMDS axis 2, were maintained throughout recovery
trajectories. More than commonly thought, post-disturbance trajectories
depended on community initial composition, that partly determined the
pool of recruited species and constrained the trajectories towards the
initial composition. The high resilience of communities taxonomy
revealed that species not belonging to the pre-disturbance community
were hardly recruited because of the commonness of dispersal limitation
among tropical tree species \citep{Svenning2005}.

Trajectories of community functional composition appeared similar in
time and close in terms of functional composition. Because
pre-disturbance surviving trees mirror the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types infrequent or absent before
disturbance. Competitive pioneers became dominant in filling the
environmental niches of high availability of light, space and nutrient
made available by the disturbance. It changed community functional
composition in the same way for all communities, towards more
resource-acquisitive strategies
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Later during the
community recovery long-lived, more resistant and shade-tolerant
species, would exclude those pioneers and restore the pre-disturbance
composition.

@@

Disturbance clearly enhanced community taxonomic richness until an
intensity threshold (20-25\% of initial AGB loss), while the taxonomic
evenness was somewhat decoupled from the disturbance intensity, as
already observed in the Guiana Shield \citep{Baraloto2012a} and in
Bornean tropical forests \citep{Cannon1998}. The IDH well predicted the
response of taxonomic richness to disturbance, and to a lesser extent
the taxonomic evenness. Post-disturbance communities combine
pre-disturbance surviving trees and recruited trees, either from the
remaining pre-disturbance community or from a disturbance-specific
community. In the Guiana Shield, the pool of true pioneer species
specifically recruited after disturbance is restricted to a few common
genera (e.g.~Cecropia spp., Vismia spp.) Below the intensity threshold
the taxonomic richness of the pre-disturbance community is maintained
and the recruitment of pioneers, previously infrequent or absent,
increases both taxonomic richness and evenness. The disturbance
intensity determined the relative importance of pioneers recruited, so
the taxonomic richness and evennes increase all the more so that
disturbance is intense. However, disturbance in the same time decreases
the taxonomic richness of the pre-disturbance community, and beyond the
intensity threshold the decrease exceeds the taxonomic enrichment from
pioneers. The taxonomic evenness for its part is less sensitive to the
loss of rare species and the disturbance impact is mitigated as pioneers
balance the hyper-dominance of undisturbed tropical forests. Conversely
the IDH was disproved for the functional diversity. The pioneer species
were functionally highly different from the pre-disturbance community
composition, so their recruitment increased the community functional
richness and eveness after disturbance whenever its intensity.

Community post-disturbance trajectories showed a first drop of the
taxonomic richness, similar among communities, followed by a species
gain determined by the disturbance intensity. Until an intensity
threshold the gain in taxonomic richness increased with the disturbance
intensity. This corresponded to the establishment of long-lived pioneers
enhancing community taxonomic richness and evenness in the long term.
Beyond the intensity threshold a few short-lived pioneers occupied the
vacated environmental space, preventing the establishment of further
species. The establishment of short-lived pioneers, functionally very
different from the pre-disturbance community, rapidly and significantly
increased the functional diversity but it started decreasing after 10
years. Below the disturbance intensity threshold on the other hand the
establishment of long-lived pionneers entailed a later and lower but
long-lasting increase of the functional diversity. Likely the
establishment of long-lasting pioneers will follow the decline of
short-live pionneers and we expect the taxonomic and functoinal
trajectories to catch up with those observed after intermediate
disturbance.

@@

The functional redundancy within the pre-disturbance functional space
decreased for 15 years because of the loss of species, and was not
compensated by the functionally different pionneers. The depth of the
redundancy decrease relied on the disturbance intensity that enhanced
the initial decrease and the functional differences of pionneers from
the pre-disturance composition. After 15 years the first established
pioneers pionneers started to be replaced by more competitive long-lived
pionneers or late-successional species resembling more the
pre-disturbance functional composition and hence increasing the
functional redundancy. This replacement though is stochastic and follows
the lottery recruitment so because of increasing competition it is
increasingly difficult recover pre-disturbance species,specifically the
most infrequent ones, as the functional redundancy recovers
\citep{Busing2002}. The recovery time proved long as for some community
it was just initiated after 30 years but it is thus difficult to
estimate \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of pre-disturbance communities and higher
chances to see the persistence of disturbance-specific communities, with
lower taxonomic richness and more pioneers
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the slowed recovery
of rare species increased the risks to loose cornerstone specie, with
unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.
